# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data

Once the file "activity.csv"" is unzipped from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip, the data are loaded into R:

```{r}
##Download file from internet, unzip and read data
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zipFile <- download.file(url, destfile = "zipFile", method="curl")  
unzipFile <- unzip("zipFile")
data <- read.csv(unzipFile)
```

## What is mean total number of steps taken per day?

The steps per day (ignoring days of all NA values for steps) are summarized by the following histogram:

```{r}
#Total up steps of input data by date
array <- tapply(data$steps, data$date, sum)
        
#Display histogram of total number of steps taken each day
hist(array, breaks=10, col = "lightblue", main = "Histogram of Steps per Day", 
     xlab = "Steps")
        
#Store mean and sum variables
meanSteps <- as.integer(mean(array, na.rm=TRUE))
medianSteps <- as.integer(median(array, na.rm=TRUE))
```

The mean number of steps taken per day (ignoring NAs) is `r meanSteps`, and the median is `r medianSteps`.

## What is the average daily activity pattern?

The following time series plot shows activity patterns from the observed data of average steps for each interval measured throughout the day:

```{r}
#Create new dataframe of average steps by interval
data$interval <- factor(data$interval)
intervalAvg <- tapply(data$steps, data$interval, mean, na.rm=TRUE)
intervalAvg <- data.frame(intervalAvg)

#Make time series plot of steps/day vs interval
plot(row.names(intervalAvg), intervalAvg$intervalAvg, type= "l")
abline(v = 835, col="red")

#Order average steps by interval in descending order
intervalAvg <- intervalAvg[order(intervalAvg, decreasing=TRUE),]
```

The maximum number of steps on average is `r intervalAvg[1]` which takes place on the interval `r names(intervalAvg[1])`, as seen in the graph above.

## Imputing missing values

Imputing the NA values is done by assuming typical activity for the corresponding intervals of days with observed data.  newData include all of these average values in lieu of NAs, but are otherwise identical to the original imported data.

```{r}
#initialize new dataframe as replica of original data
newData <- data

#transform averages back to dataframe to reference for replacement data:
intervalAvg <- data.frame(intervalAvg)

#Replace means into newData if steps measurements are NA
for (i in seq_len(nrow(newData))) {
        if (is.na(newData[i, "steps"])) {
                newData[i, "steps"] <- 
                intervalAvg[row.names(intervalAvg) == newData[i,"interval"],]
                } 
        }
```

The new data with imputed values are summarized with the same code as the original:

```{r}
#Total up steps of input data by date
array <- tapply(newData$steps, newData$date, sum)
        
#Display histogram of total number of steps taken each day
hist(array, breaks=10, col = "lightblue", main = "Histogram of Total   
     Steps per Day (imputed)", xlab = "Steps")
        
#Create array summary and store mean, median, and sum variables
meanSteps <- as.integer(mean(array, na.rm=TRUE))
medianSteps <- as.integer(median(array, na.rm=TRUE))

```
It's clear that the measures of central tendency have not changed (much), as the mean is still `r meanSteps` and the median has only changed to `r medianSteps`.  The impact of imputing missing data has added more observations right around the mean, which can be expected since the imputed are averages themselves.

## Are there differences in activity patterns between weekdays and weekends?

New imputed dataset is factored into weekday/weekend observations and plotted  based on those factors:

```{r}
 ##change class to date, assign weekdays
newData$date <- as.Date(newData$date)

#create new factor variable of day type
newData$dayType <- weekdays(newData$date)

require(car)
newData$dayType <- recode(newData$dayType, "c('Monday', 'Tuesday', 'Wednesday'                
                , 'Thursday', 'Friday') = 'weekday';
                c('Saturday', 'Sunday') = 'weekend' ")
newData$dayType <- factor(newData$dayType)
        
#melt and cast
require(reshape2)
newMelt <- melt(newData, id=c("interval", "dayType"), measure.vars="steps" )
newCast <- dcast(newMelt, interval + dayType ~ variable, mean)
        
require(lattice)
xyplot(steps ~ interval | dayType, data=newCast, type = "l", layout = c(1,2))
```
